---
title: "fetal_adult_analysis.Rmd"
output: html_document

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pheatmap)
library(tidyverse)
library(ggplot2)
library(DESeq2)
library(dplyr)
library(RColorBrewer)
my_pal <- brewer.pal(n = 10, name = 'RdBu')
```


```{r - results from pipeline}
#read in tpm
tpm <- read.table("/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/results/star_salmon/salmon.merged.gene_tpm.tsv", header = T)



#read in counts
counts <- read.table("/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/results/star_salmon/salmon.merged.gene_counts.tsv", header = T)

gene_name <- counts %>%
  select(gene_id, gene_name)

counts <- counts %>%
  column_to_rownames(var = "gene_id") %>%
  select(-gene_name)

#clean those with all 0 counts
counts <- counts[rowSums(counts) > 0,]

```

```{r - samples}
# Let's set names for 2 distinct analysis:
#   1) all fetal tissues as replicates (9) for : fetal
#   2) each fetal stage as a group, 3 replicates each

samplesheet <- read.csv("/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/samplesheet.csv")

tissue <- c("fetal", "fetal","fetal","fetal","fetal","fetal","fetal","fetal","fetal", "adult", "adult", "adult", "adult")

samples <- samplesheet %>%
  select(sample) %>%
  mutate(tissue = tissue)

samples$tissue_time <- gsub("W_[1-9]", "W", samples$sample)
samples$tissue_time <- gsub("_heart_1[0-3]", "", samples$tissue_time)

#transform comparisons in factor for deseq
factor_1 <- c("fetal", "adult")
factor_2 <- c("fetal_9W", "fetal_12W", "fetal_16W", "adult")

samples$tissue <- factor(samples$tissue, levels = factor_1 )
samples$tissue_time <- factor(samples$tissue_time, levels = factor_2 )
```

```{r - calculate mean tpm per sample}
# Calculate mean following two type of sample grouping:
#   1) all fetal tissues as replicates (9) for : fetal
#   2) each fetal stage as a group, 3 replicates each

tpm_mean_adult_fetal <- tpm %>%
  pivot_longer(cols = adult_heart_10:fetal_9W_9, names_to = "sample", values_to = "tpm") %>%
  left_join(samples) %>%
  group_by(gene_id, tissue) %>% ####mean by tissue
  mutate(mean_tpm_tissue = mean(tpm)) %>%
  mutate(sd_tissue = sd(tpm, na.rm=TRUE))

tpm_mean_adult_fetal <- tpm_mean_adult_fetal %>%
  ungroup() %>%
  group_by(gene_id, tissue_time) %>% ###mean by tissue_time
  mutate(mean_tpm_time = mean(tpm)) %>%
  mutate(sd_time = sd(tpm, na.rm=TRUE))

write_rds(tpm_mean_adult_fetal, "/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/tpm_mean.rds")
```

```{r - prep for deseq}
#reordering and checking sample_id (it MUST be the same order in both)
samples$sample <- as.character(samples$sample)

col_order <- c(samples$sample)
counts <- counts[, col_order]
all(colnames(counts) == samples$sample)

#change to matrix
counts <- as.matrix(counts)

# Now we need to round the values and change to integer from numeric.
counts <- round(counts)

# change counts to integer (this is required by DESeq)
mode(counts) <- "integer"
```


# CARDEL : ENSG00000260802


Let's run the DE analysis:

```{r}
# Analysis 1: fetal vs adult

dds <- DESeqDataSetFromMatrix(counts, samples,
                              design = ~ tissue)
dds_run <- DESeq(dds)
resultsNames(dds_run)

#save analysis
write_rds(dds_run, "/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/dds_run1.rds")

adult_fetal <- as.data.frame(results(dds_run, name = "tissue_adult_vs_fetal")) %>%
  rownames_to_column(var = "gene_id") %>%
  left_join(gene_name)

#save table
write.csv(adult_fetal, "/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/adult_fetal_deseq_results.csv")

# extract rlog_counts if want to plot heatmaps
rlog_counts <- dds %>%
  rlog() %>%
  assay() %>%
  as.data.frame() 

#save rlog_counts
write_rds(rlog_counts,"/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/adult_fetal_rlog_counts.csv" )

```

```{r}
# Analysis 2: fetal timecourse vs adult
dds <- DESeqDataSetFromMatrix(counts, samples,
                              design = ~ tissue_time)
dds_run <- DESeq(dds)
resultsNames(dds_run)

#save analysis
write_rds(dds_run, "/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/dds_run2.rds")

# extract rlog_counts if want to plot heatmaps
rlog_counts_timecourse <- dds %>%
  rlog() %>%
  assay() %>%
  as.data.frame() 

#save rlog_counts
write_rds(rlog_counts_timecourse,"/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/timecourse_rlog_counts.csv" )


# extract all results

timecourse_adult_16W <- as.data.frame(results(dds_run, name = "tissue_time_adult_vs_fetal_16W")) %>%
  rownames_to_column(var = "gene_id") %>%
  left_join(gene_name) %>%
  mutate(comparison = "tissue_time_adult_vs_fetal_16W")

#save table ##change for each comparison
write.csv(timecourse_adult_12W, "/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/timecourse_adult_16W_deseq_results.csv")

# use this to change level:
dds$tissue_time <- relevel(dds$tissue_time, ref = "fetal_16W")
dds_run <- DESeq(dds)
resultsNames(dds_run)
# go back above to write results

```

```{r}

save(samples, tpm_mean_adult_fetal, gene_name, rlog_counts, adult_fetal, rlog_counts_timecourse, timecourse_12W_9W, timecourse_16W_9W, timecourse_adult_9W, timecourse_16W_12W, timecourse_adult_12W, timecourse_adult_16W,
     file = "/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/adult_fetal.RData")

load("/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/adult_fetal.RData")
```

# CARDEL : ENSG00000260802

### Apply filters to find those DEG

```{r - set cutoff}
padj_cutoff <- 0.05
logFC <- 2 # FC = 4  #trying to be conservative 
```


```{r}
# analysis 1:

# all DEGs
adult_fetal_filtered <- adult_fetal %>%
  filter(padj <= padj_cutoff) %>%
  filter(log2FoldChange >= logFC | log2FoldChange <= -logFC)

# the goal is filter those up in fetal tissue

fetal_enriched <- adult_fetal_filtered %>%
  filter(log2FoldChange <= -logFC)

### CARDEL is here! enriched in fetal vs adult

# those which would be excluded because are up in adult:
adult_enriched <- adult_fetal_filtered %>%
  filter(log2FoldChange >= logFC)
```

```{r}
# analysis 2:

# following the timecourse, two patters:
#   A) increase during fetal e decrease in adult (CARDEL increase, but not sig)
#   B) decrease in fetal and more in adult
#   C) increase fetal e continue increasing in adult - not our goal, but will be used as filter

#A)
timecourse_12W_9W_filter <- timecourse_12W_9W %>%
  filter(padj <= padj) %>%
  filter(log2FoldChange >= logFC)

timecourse_16W_12W_filter <- timecourse_16W_12W %>%
  filter(padj <= padj) %>%
  filter(log2FoldChange >= logFC)

timecourse_adult_16W_filter <- timecourse_adult_16W %>%
  filter(padj <= padj) %>%
  filter(log2FoldChange <= -logFC) ### decreases in adult
  ## this is the only that CARDEL is - meaning that it doesn't increase in this fetal timecourse, but is sig in adults vs 16w


#C)
timecourse_12W_9W_filter <- timecourse_12W_9W %>%
  filter(padj <= padj) %>%
  filter(log2FoldChange >= logFC)

timecourse_16W_12W_filter <- timecourse_16W_12W %>%
  filter(padj <= padj) %>%
  filter(log2FoldChange >= logFC)

timecourse_adult_16W_filterC <- timecourse_adult_16W %>%
  filter(padj <= padj) %>%
  filter(log2FoldChange >= logFC) # keep increasing in adult

increase_all <- intersect(intersect(timecourse_12W_9W_filter$gene_id, timecourse_16W_12W_filter$gene_id), timecourse_adult_16W_filterC$gene_id)

both <- intersect(increase_all, adult_enriched$gene_id)
#all are represented there, so let's only use the adult vs fetal analysis (1)
```

```{r - plots - NADA FICOU BOM, APAGAR}
rlog_counts_id <- rlog_counts %>%
  rownames_to_column(var = "gene_id")

# Let's plot a heatmap of all fetal enriched - all replicates
fetal_enriched_list <- fetal_enriched %>%
  select(gene_id) %>%
  left_join(rlog_counts_id) %>%
  column_to_rownames(var = "gene_id")


fetal_enriched_list_scaled <- t(scale(t(fetal_enriched_list))) 

pheatmap::pheatmap(fetal_enriched_list_scaled, show_rownames = F, 
                   cluster_rows = F, cluster_cols = T,
                   breaks = seq(-3, 3, length.out = length(my_pal)),
                   color = rev(my_pal), border_color = NA, fontsize_row = 8)

# How it looks using  (log) mean tpm ====  uma bosta, tentei

# How it looks using rlog_mean === ficou a mesma bosta do mean tpm
rlog_mean <- rlog_counts_id %>%
  pivot_longer(!gene_id, names_to = "sample", values_to = "rlog_count") %>%
  left_join(samples) %>%
  group_by(gene_id, tissue) %>% ####mean by tissue
  mutate(rlog_mean = mean(rlog_count)) %>%
  select(gene_id, tissue, rlog_mean) %>%
  distinct() %>%
  pivot_wider(names_from = tissue, values_from = rlog_mean)


fetal_enriched_list <- fetal_enriched %>%
  select(gene_id) %>%
  left_join(rlog_mean) %>%
  column_to_rownames(var = "gene_id")

fetal_enriched_list_scaled <- t(scale(t(fetal_enriched_list))) 
pheatmap::pheatmap(fetal_enriched_list_scaled, show_rownames = F, 
                   cluster_rows = F, cluster_cols = T,
                   breaks = seq(-0.5, 0.5, length.out = length(my_pal)),
                  color = rev(my_pal), border_color = NA, fontsize_row = 8)

```

## Comparing with hESC diff

```{r}
# First, let's re-analyze the D4 x D9 data
# this dataset is already filtered by FDR (padj) < 0.05
# from old analysis ==== watch for update names etc #TODO: re-run nextflow for these <<<<<<<<<<<<<

d4_d9 <- read.csv("/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/results/DEG_D4_D9_FDR0.05.csv")

up_d9 <- d4_d9 %>%
  filter(logFC >= 2) #keep same parameter than earlier, but maybe try others
```

```{r}
# Now, we can compare those up in D9 and fetal_enriched == these will be our targets

#keep only gene_id_stable
fetal_enriched$gene_id_stable <- gsub("[.]\\d+", "", fetal_enriched$gene_id)
up_d9 <- up_d9 %>%
  dplyr::rename(gene_id_stable = Row.names)


targets <- up_d9 %>%
  filter(gene_id_stable %in% fetal_enriched$gene_id_stable)
#98

#bring updated names ===== temp
gene_name$gene_id_stable <- gene_id_stable <- gsub("[.]\\d+", "", gene_name$gene_id)

targets <- targets %>%
  left_join(gene_name)

write.csv(targets, "/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/d9_fetal_targets.csv")
```

# WHY CARDEL?
# CARDEL : ENSG00000260802

Now is the challenge of finding a nice story to tell about CARDEL

```{r}
# exclude those that are annotated in BP GOs:

bp_go <- read.csv("/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/gProfiler_hsapiens_8-6-2022_7-26-48 PM__intersections_BP.csv")

bp_go_genes <- bp_go %>%
  filter(source == "GO:BP") %>%
  mutate(genes_name = strsplit(as.character(intersections), ",")) %>%
  unnest(genes_name) %>%
  select(genes_name) %>%
  distinct()

targets_filter_bp <- targets %>%
  filter(!gene_name %in% bp_go_genes$genes_name)
#good, list of 37

#rank by logFC d4xd9 AND opposite in fetal_enriched
fetal_logFC <- fetal_enriched %>%
  select(gene_id_stable, log2FoldChange)

targets_filter_bp <- targets_filter_bp %>%
  left_join(fetal_logFC)

top_targets <- targets_filter_bp %>%
  slice_max(logFC, n = 10)

#let's take a look in the tpm from tissue
targets_tpm <- tpm_mean_adult_fetal %>%
  filter(gene_id %in% targets_filter_bp$gene_id)

```

NICE! So far we can justify CARDEL by:
- among higher logFC d4 x d9
- among the highest RPKM in D9
- it is not any weird stuff like pseudogene or not interesting like antisense, sense-over (maybe say independent locus)

```{r - save space}
save(adult_fetal, fetal_enriched, adult_enriched, d4_d9, up_d9, targets, targets_filter_bp,
     file = "/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/adult_fetal_d4_d9_results.RData")
```

