---
title: "plots"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(DESeq2)
library(VennDiagram)
library(ggdendro)
library(GenomicRanges)
library(RColorBrewer)
library(ggrepel)
source("/scratch/Shares/rinn/isabela/rna_protein_complexes/analysis/util/_plot_theme.R")

my_pal <- brewer.pal(n = 11, name = 'RdBu')
```


```{r - color palette}
display.brewer.pal(n = 11, name = 'RdBu') #11 max colors in the palette

brewer.pal(n = 11, name = 'RdBu') #gives the color codes
#[1] (red) "#67001F" "#B2182B" "#D6604D" "#F4A582" "#FDDBC7" "#F7F7F7" "#D1E5F0" "#92C5DE" "#4393C3" #"#2166AC"
#[11] "#053061" (blue)
my_pal <- brewer.pal(n = 11, name = 'RdBu')

#use this in ggplot
+ scale_color_brewer(palette = "RdBu") +
```

For color palette https://www.datanovia.com/en/blog/the-a-z-of-rcolorbrewer-palette/

```{r}
load("/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/adult_fetal.RData")
load("/scratch/Shares/rinn/isabela/CARDEL/fetal_adult/analysis/adult_fetal_d4_d9_results.RData")
```

##working here
```{r - volcano plot}

#D4_D9 <<<<<<<<<<<<<<<<<<<<<<<<<<<<< usar aqui a nova table

D4_D9_volcano <- d4_d9 %>%
  select(Row.names, log2FoldChange, padj, external_gene_name)


#adult_fetal
# padj < 0.05, all DEG

fetal_volcano <- adult_fetal %>%
  select(gene_id, gene_name, log2FoldChange, padj)

#clean NA from padj
fetal_volcano <- fetal_volcano[complete.cases(fetal_volcano), ]

#assign UP & DOWN
fetal_volcano$deg <- "NO" #set non-deg first
fetal_volcano$deg[fetal_volcano$log2FoldChange >= 2 & fetal_volcano$padj <= 0.05] <- "UP"
fetal_volcano$deg[fetal_volcano$log2FoldChange <= -2 & fetal_volcano$padj <= 0.05] <- "DOWN"

#label for top 10 genes names <<<<<<<<<<<<<<<<< name here some of 98<< need work
top_padj <- deg10_volcano %>%
  filter(padj <= 0.05)
top_up <- top_padj[with(top_padj,order(-log2FoldChange)),]
top_up <- top_up[1:10,]

top_down <- top_padj[with(top_padj,order(log2FoldChange)),]
top_down <- top_down[1:10,]

top_names <- rbind(top_down, top_up)
top_names <- top_names %>%
  mutate(label = top_names$external_gene_name)

deg10_volcano <- deg10_volcano %>%
  left_join(top_names)

#aes(label = label)

#VOLCANO PLOT = log2FC & -log10(padj)

ggplot(fetal_volcano, aes(x=log2FoldChange, y=-log10(padj), col=deg)) + 
  geom_point() +
  #theme_minimal() +
  scale_color_manual(values=c("#053061", "black", "#67001F")) + #color assigned in alphabetical order (use same as heatmap = brewer palette RdBu)
  #geom_text_repel() + #this add names of top10 genes
  theme(text = element_text(size = 12), axis.text = element_text(size = 12))  

#save as pdf landscape 4 x 3.5
